---
title: "Intra_Differential_Expression"
output: html_document
---

# Read-In Libraries

```{r, include =FALSE}
library(tidyverse)
library(glue)
library(devtools)
library(vroom)
library(GEOquery)
library(biomaRt)
library(dplyr)
library(org.Hs.eg.db)
library(DESeq2)
library(clusterProfiler)
library(RNAseqQC)
library(edgeR)
source("Project_1_function_lab.R")
```

```{r}
covid19_ds1 <- readRDS("./post_processed_data/gse202805_covid19_ds1.rds")
covid19_ds2 <- readRDS("./post_processed_data/gse152641_covid19_ds2.rds")
covid19_ds4 <- readRDS("./post_processed_data/gse227116_covid19_ds4.rds")
covid19_ds124 <- readRDS("./post_processed_data/covid_dds_124.rds")


# Adding DataSet Moniceurs
colData(covid19_ds1)$ds_origin <- "GSE202805"
colData(covid19_ds2)$ds_origin <- "GSE152641"
colData(covid19_ds4)$ds_origin <- "GSE227116"
```

<https://olvtools.com/en/documents/edger>

# Part2_Project 1 Intra Diffferential Expression of the DataSets

### Perform EdgeR Differential Expression and plot a volcano plot for each condition comparison within the DDS object

```{r}

#Arguments
dds_object <- covid19_ds1
condition_col = "condition"
ref_group_name = "healthy"


#Performs inital differential expression on your DDS Object's raw counts dataframe
# The use give the condition column from the metadata to make the test parameters
# and the control group
results_dds1 <- edgeR_diffexp(covid19_ds1, condition_col,  ref_group_name)
results_dds2 <- edgeR_diffexp(covid19_ds2, condition_col, "Healthy")
results_dds4 <- edgeR_diffexp(covid19_ds4, condition_col, ref_group_name)
```

### Filtering Differential expression results by differential expression and pvalue

```{r}
require(glue)


# function starts here
#create a list of arg
dds_object <- covid19_ds2

edgeR_results <- results_dds2

#controls the intial filtering of DEGs directly after the edgeR differential expression
deg_log2fc_thresh <- 1

#controls the intial p.val thresh of DEGs directly after the edgeR differential expression
deg_pval_adj_thresh <- 0.05

# controls the volcano plot's lower x-axis limit on the image
vol_plot_lower_xlim <- -10

#controls the volcano plot's upper x-axis limit on the image
volcano_plot_upper_xlim <- 10

# ticker step of the plot on both x and y axis
cartesian_step <- 5

#over-representation test pvalue threshold
ora_pval_adj_threshold <- 0.05


#save vector the length of the edgeR results with comparisons
diffexp_results <- vector("list", length(edgeR_results))
names(diffexp_results) <- names(edgeR_results)

# for each comparison
for (i in seq_along(edgeR_results)) {

  volcano_res <- volcano_plot(
    edgeR_results[[i]],
    log2fc_thresh = deg_log2fc_thresh,
    p.val_adj_thresh = deg_pval_adj_thresh,
    lower_xlim = vol_plot_lower_xlim,
    upper_xlim = volcano_plot_upper_xlim,
    step = cartesian_step
  )
  #for each comparison within the edgeR diff exp analysis
  # this function will save the filtered DEG table
  
  #save directly to the initialized vector
  diffexp_results[[i]] <- list(
    #saving the first subcript content of the volcano_res function
    data = volcano_res[[1]],

    vol_plot = volcano_res[[2]] +
      labs(title = glue::glue(
        "Volcano Plot: {names(edgeR_results)[i]}"
      ))

  )
}

# save the filtered Differential Expression Results
filtered_results <- lapply(diffexp_results , function(x) x$data)

#report for each comparison the number of up and down regulated DEGS
for(i in seq_along(filtered_results)){
  
  num_sig_up_degs <- unique(filtered_results[[i]]$geneid[filtered_results[[i]]$log2foldchange > deg_log2fc_thresh & filtered_results[[i]]$p.val_adj < deg_pval_adj_thresh])
  
  num_sig_down_degs <-unique(filtered_results[[i]]$geneid[filtered_results[[i]]$log2foldchange < -deg_log2fc_thresh & filtered_results[[i]]$p.val_adj < deg_pval_adj_thresh])
  
  total <- length(num_sig_down_degs) + length(num_sig_up_degs)
  
  message(
    glue(
      "Within Dataset {unique(colData(dds_object)$ds_origin)}, ",
      "Comparison: {names(filtered_results)[i]} has ",
      "{length(num_sig_up_degs)} significantly up-regulated DEGs and ",
      "{length(num_sig_down_degs)} down-regulated DEGs",
      "Making a total of {total} DEGS"
    )
  )
}


#Plot the Heatmap of the Differential Expression Anaalysis
heatmap_visuals <- heatmap_function(dds_object, filtered_results)


#saving visuals for Volcano plot
results_vol_plots <- lapply(diffexp_results, function(x) x$vol_plot)


#perform an EnrichGO ORA
#ORA only takes in significant DEGs Answers the question:
#Which pathways are overrepresented in my list of significant genes? 

ora_output_up <- ora_enrichgo(
  # Output of the differential expression results.
  filtered_results,
  # Which Direction of the DEGs (Upregulated("up")/Down("down"))
  direction = "up",
  #How strict do you want your DEGs entered into the ORA?
  log2fc_thresh = deg_log2fc_thresh,
   # what is the pval_thresh of the ORA
  pval_thresh = ora_pval_adj_threshold,
  #How many terms do you wantv visualized in your dotplots?
  top_terms = 15
)

ora_output_down <- ora_enrichgo(
  # Output of the differential expression results.
  filtered_results,
  # Which Direction of the DEGs (Upregulated("up")/Down("down"))
  direction = "down",
  #How strict do you want your DEGs entered into the ORA?
  log2fc_thresh = deg_log2fc_thresh,
  # what is the pval_thresh of the ORA
  pval_thresh = ora_pval_adj_threshold,
  #How many terms do you wantv visualized in your dotplots?
  top_terms = 15
)

#Unpacking ORA from enrichGO

#Unpacking ORA from GO and align to your differential expression results for the input condition

#ORA only takes in significant DEGs Answers the question:

#Which pathways are overrepresented in my list of significant genes?


# making a data table for each Description Term that is consider statistically significant within the context of the ORA

significant_degs <- filtered_results

key_down <- ora_output_down$enrichgo_results

key_up <-ora_output_up$enrichgo_results
  

unpacked_results_down <-  enrichgo_unpack_ver3(initial_table_list = significant_degs,
                                                 key = key_down,
                                                 pval_adj_threshold = ora_pval_adj_threshold,
                                                 convert = FALSE,
                                                 from_type = "ensembl",
                                                 to_type = "symbol",
                                                 ensembl_dataset = "hsapiens_gene_ensembl")

unpacked_results_up <-  enrichgo_unpack_ver3(initial_table_list = significant_degs,
                                                 key = key_up,
                                                 pval_adj_threshold = ora_pval_adj_threshold,
                                                 convert = FALSE,
                                                 from_type = "ensembl",
                                                 to_type = "symbol",
                                                 ensembl_dataset = "hsapiens_gene_ensembl")
```

### Displaying All the Results

```{r}
diffexp_results
filtered_results
heatmap_visuals
ora_output_up
ora_output_down
```

### **Covid-19 DS1 GSE202805 significant DEGs abs(log2fc) \> 1 and pval_adj \< 0.05**

**healthy-convalalescent: \# DEGs total = 114**

Upregulated: 19

Down Regulated: 95

**Healthy-covid19: #DEGs total = 295**

UpRegulated: 61

Downregulated: 234

Convalescent-covid19: \# DEGs total = 1408

Upregulated: 535 Sig up

Downregulated: 87

### Covid-19 DS2 GSE152641 **significant DEGs abs(log2fc) \> 1 and pval_adj \< 0.05**

**Healthy-covid19: #DEGs total = 591**

UpRegulated: 112 sig up

Downregulated: 479 sig down

# Healthy vs COVID-19 Comparisons

Covid-19 Dataset 1 (GSE202805) top over-represented pathways by p-values of hypergeometric test

```{r}
df1_ds1 <- unpacked_results_down$`healthy-covid19`$`immunoglobulin mediated immune response`
df2_ds1 <- unpacked_results_down$`healthy-covid19`$`B cell mediated immunity`
df3_ds1 <- unpacked_results_down$`healthy-covid19`$`antibacterial humoral response`
df4_ds1 <- unpacked_results_down$`healthy-covid19`$`mucosal immune response`
df5_ds1 <- unpacked_results_down$`healthy-covid19`$`antimicrobial humoral response`

df_ora_filtered <- unique(bind_rows(df1_ds1, df2_ds1, df3_ds1, df4_ds1, df5_ds1), by = "gene_id")

df_ora_filtered <- list(df_ora_filtered)
names(df_ora_filtered) <- 'healthy-covid19'

heatmap_function(dds_object, df_ora_filtered)
```

```{r, include=FALSE}
df1_ds2 <- unpacked_results_down$`Healthy-COVID19`$`defense response to virus`
df1_ds2_ora_filtered <- unique(bind_rows(df1_ds2), by = "gene_id")

df1_ds2_ora_filtered <- list(df1_ds2_ora_filtered)
names(df1_ds2_ora_filtered) <- 'healthy-covid19'

heatmap_function(dds_object, df1_ds2_ora_filtered)
```

# Healthy vs Convalescent Comparisons

(per Dataset)

Covid-19 DS1 has 131 unique DEGs total for this comparison

# Covid-19 vs Convalescent Comparison

(per Dataset)

Covid-19 DS1 had 2303 unique signifcant (abs(log2FC) \>= 1, pval \< 0.05)DEGs total for this comparison

# Intergrated Dataset Differential Expression

```{r}
unique(covid19_ds124@colData$condition)
```

```{r}
dds_object <- covid19_ds124 
condition_col = "condition"
ref_group_name = "healthy"
results_dds124 <- edgeR_diffexp(covid19_ds124,
                                condition_col,
                                ref_group_name)
```

```{r, include=FALSE}

# 1. Initialize dataset-specific starting table
initial_table_124 <- covid19_ds124

# 2. Set up the list container for raw results
diffexp_results_124 <- vector("list", length(results_dds124))
names(diffexp_results_124) <- names(results_dds124)

# 3. Process each comparison through the volcano function
for (i in seq_along(results_dds124)) {

  volcano_res <- volcano_plot(
    results_dds124[[i]],
    log2fc_thresh = 1,
    p.val_thresh = 0.01,
    lower_xlim = -10,
    upper_xlim = 10,
    step = 5
  )
  
  # Store both the data frame and the customized plot
  diffexp_results_124[[i]] <- list(
    data = volcano_res[[1]],
    vol_plot = volcano_res[[2]] +
      labs(title = glue::glue("Volcano Plot: {names(results_dds124)[i]}"))
  )
}

# 4. Add the Datatables within a seperate variable for future parsing
filtered_results_124 <- lapply(diffexp_results_124, function(x) x$data)

# 5. Generating Heatmap for each comparison
heatmap_visuals_124 <- heatmap_function(initial_table_124, filtered_results_124)
```

```{r}
#saving visuals for Volcano plot
results_vol_plots <- lapply(diffexp_results_124, function(x) x$vol_plot)


#perform an EnrichGO ORA
#ORA only takes in significant DEGs Answers the question:
#Which pathways are overrepresented in my list of significant genes? 

ora_output_up_124 <- ora_enrichgo(
  # Output of the differential expression results.
  filtered_results_124,
  # Which Direction of the DEGs (Upregulated("up")/Down("down"))
  direction = "up",
  #How strict do you want your DEGs entered into the ORA?
  log2fc_thresh = 1,
   # what is the pval_thresh of the ORA
  pval_thresh = 0.05,
  #How many terms do you wantv visualized in your dotplots?
  top_terms = 15
)

ora_output_down_124 <- ora_enrichgo(
  # Output of the differential expression results.
  filtered_results_124,
  # Which Direction of the DEGs (Upregulated("up")/Down("down"))
  direction = "down",
  #How strict do you want your DEGs entered into the ORA?
  log2fc_thresh = 1,
  # what is the pval_thresh of the ORA
  pval_thresh = 0.05,
  #How many terms do you wantv visualized in your dotplots?
  top_terms = 15
)
```

```{r}
# making a data table for each Description Term that is consider statistically significant within the context of the ORA

initial_tables <- filtered_results_124

key_down <- ora_output_down_124$enrichgo_results

key_up <-ora_output_up_124$enrichgo_results
  
pval_adj_threshold <- 0.001

unpacked_results_down <-  enrichgo_unpack_ver3(initial_table_list = initial_tables,
                                                 key = key_down,
                                                 pval_adj_threshold = 0.005,
                                                 convert = FALSE,
                                                 from_type = "ensembl",
                                                 to_type = "symbol",
                                                 ensembl_dataset = "hsapiens_gene_ensembl")

unpacked_results_up <-  enrichgo_unpack_ver3(initial_table_list = initial_tables,
                                                 key = key_up,
                                                 pval_adj_threshold = 0.005,
                                                 convert = FALSE,
                                                 from_type = "ensembl",
                                                 to_type = "symbol",
                                                 ensembl_dataset = "hsapiens_gene_ensembl")

```

```{r}
heatmap_visuals_124

heatmap_folder <- "heatmap_images"


if(!dir.exists(heatmap_folder)) dir.create(heatmap_folder, recursive = TRUE)


for(h in seq_along(heatmap_visuals_124)){
gg_patchwork(heatmap_visuals_124[[h]], 
             glue("{heatmap_folder}/{names(heatmap_visuals_124)[h]}.png"),
             width = 10,
             height = 20)
}
```

```{r}
results_vol_plots
```

```{r}
ora_output_up_124$enrichgo_visuals
ora_output_down_124$enirchgo_visuals

```

```{r}

ora_output_down_124 <- ora_enrichgo(
  # Output of the differential expression results.
  filtered_results_124,
  # Which Direction of the DEGs (Upregulated("up")/Down("down"))
  direction = "down",
  #How strict do you want your DEGs entered into the ORA?
  log2fc_thresh = 1,
  # what is the pval_thresh of the ORA
  pval_thresh = 0.05,
  #How many terms do you wantv visualized in your dotplots?
  top_terms = 15
)
```

```{r}
ora_output_down_124
```
