---
title: "Project_1"
output: html_document
---

Source Challenge: <https://github.com/mitreac/BIDS2026_projects>

```{r}
library(tidyverse)
library(devtools)
library(GEOquery)
library(vroom)
library(biomaRt)
library(dplyr)
library(org.Hs.eg.db)
library(DESeq2)
library(clusterProfiler)
library(RNAseqQC)
```

## Reading in GSE202805 file

80 Samples:

Platform: Illumina HiSeq 4000 (Homo sapiens)

Case Groups: 4 acute-Mild, 6 Acute Moderate, 32 Acute-Severe, 28 Covalescent

Control Groups: 10 healthy controls

Total:

```{r}
gse_202805 <- getGEO("GSE202805", GSEMatrix = TRUE)

gse_202805 <- gse_202805[[1]]

expr_202805 <- exprs(gse_202805)

meta_202805 <- pData(gse_202805)

head(expr_202805)
head(meta_202805)

getGEOSuppFiles("GSE202805")
```

```{r}
counts <- read.delim(gzfile("GSE202805_genecounts_20220406.txt.gz"))

colnames(counts) <- sub("\\.bam$", "", colnames(counts))

counts <- counts %>% dplyr::select(-c(Chr, Start, End, Strand, Length))


meta_data <- meta_202805
meta_data <- meta_data %>%
  rownames_to_column(var = "samples") %>%
  dplyr::select(samples, title, characteristics_ch1) %>%
  mutate(characteristics_ch1 = gsub("group:", "", characteristics_ch1)) %>%
  dplyr::select(-c(samples))

mapping <- c(meta_data$characteristics_ch1)
names(mapping) <- meta_data$title

colnames(counts) <- mapping[colnames(counts)]

colnames(counts) <- paste(names(mapping), colnames(counts), sep = "_")

colnames(counts)[1] <- "geneid"


# Character Specific Clean Up
counts$geneid <- gsub("\\..*", "", counts$geneid)
colnames(counts) <- gsub(" ", "", colnames(counts))


# MetaData construction
metadata <- data.frame(
  sample = colnames(counts),
  treatment = dplyr::case_when(
    grepl("_acute$", colnames(counts)) ~ "acute",
    grepl("_convalescent$", colnames(counts)) ~ "convalescent",
    grepl("healthy$", colnames(counts)) ~ "healthy",
    TRUE ~ NA_character_
  )
) %>% slice(., 2:nrow(.))
```

```{r}
# summarizing replicated genes across all samples
counts <- counts %>%
  group_by(geneid) %>%
  summarise(across(where(is.numeric), sum))

counts <- counts %>% column_to_rownames(var = "geneid")

```

```{r}
counts
```

```{r}
metadata
```

```{r}
covid19_ds1 <- make_dds(counts = counts , metadata = metadata, ah_record = "AH89426")
```

```{r}
plot_total_counts(covid19_ds1)
plot_library_complexity(covid19_ds1)
plot_gene_detection(covid19_ds1)
plot_biotypes(covid19_ds1)
```

Variance Stabilization

```{r}
mean_sd_plot(covid19_ds1)
vsd_ds1 <- vst(covid19_ds1)
mean_sd_plot(vsd_ds1)
```

Filtering out low count Genes

```{r}
covid19_ds1 <- filter_genes(covid19_ds1, min_count = 5, min_rep = 4)
```

## Reading in GSE227116

65 blood Samples recoved form 46 COVID-19 convalescent patients and 10 healthy donors were analyzed by RNA- Sequencing

Platform: Illumina HiSeq 2500 (Homo sapiens)

10 healthd donors, 65 blood samples.

```{r}
gse_227116 <- getGEO("GSE227116", GSEMatrix = TRUE)
gse_227116 <- gse_227116[[1]]

expr_227116 <- exprs(gse_227116)
meta_227116 <- pData(gse_227116)

head(expr_227116)
head(meta_227116)

getGEOSuppFiles("GSE227116")
```

```{r}
meta_227116
```

```{r}
urld <- "https://www.ncbi.nlm.nih.gov/geo/download/?format=file&type=rnaseq_counts"
path <- paste(urld, "acc=GSE227116", "file=GSE227116_raw_counts_GRCh38.p13_NCBI.tsv.gz", sep="&");
tbl <- as.matrix(data.table::fread(path, header=T, colClasses="integer"), rownames=1) %>% as.data.frame()
tbl
```

```{r}

# Cleaning the Text and shuffling columns around
meta_227116$samples <- row.names(meta_227116)

rownames(meta_227116) <- NULL

names(meta_227116)[colnames(meta_227116) == "characteristics_ch1"] <- 'condition'

meta_227116 <- meta_227116 %>%
  dplyr::select(samples, condition, title) %>%
  dplyr::mutate(condition = gsub("^status: ", "", condition))


meta_227116 <- meta_227116 %>% 
            dplyr::mutate(condition = gsub("COVID-19", "", condition)) %>% 
            dplyr::mutate(condition = trimws(condition)) %>%
            dplyr::mutate(condition = gsub("s$","", condition))


#Mapping sample to the Condition
mapping <- c(meta_227116$condition)
names(mapping) <- meta_227116$samples

column_conditions <- mapping[colnames(tbl)]

new_names <- paste(colnames(tbl), column_conditions, sep = "_")

colnames(tbl) <- new_names

tbl <- rownames_to_column(tbl, "geneid")

# Summarizing the repeated geneids
tbl <- tbl%>%
  group_by(geneid) %>%
  summarise(across(where(is.numeric), sum))

nrow(tbl)
# converting entrezids  to ensemblids


# Iterative conversion function
# arguments:input, from, to, iterations, species

#managig RAM Space on a 16Gb vector limit 
rm(expr_227116, gse_227116)
```

# Total genes ids after summarization and concatentation of gene counts : 39,376

function layers

1.  mapIds from AnnotationDbi
2.  getBM from cluster Profiler
3.  bitr from clusterProfiler
4.  gconvert from g:Profiler

```{r}


# initial args
from_type <- "ENTREZ"

to_type <- "ENSEMBL"

org_db <- org.Hs.eg.db

# Harmoizing between each of the available annotation functions
harmonizer <- list(
    ENTREZ = list(local = "ENTREZID", biomart = "entrezgene_id", gconvert = "ENTREZGENE_ACC"),
    ENSEMBL = list(local = "ENSEMBL",  biomart = "ensembl_gene_id", gconvert = "ENSG"),
    SYMBOL  = list(local = "SYMBOL",   biomart = "external_gene_name",
                   gconvert = "SYMBOL")
  )

local_from   <- harmonizer[[from_type]]$local
local_to     <- harmonizer[[to_type]]$local
  
biomart_from <- harmonizer[[from_type]]$biomart
biomart_to   <- harmonizer[[to_type]]$biomart


# enter mapping
input_ids <- tbl$geneid

unique_ids <- unique(input_ids)


#Intialize final results
unique_ids_len <- length(unique_ids)

final_results <- c(replicate(unique_ids_len, NA))

names(final_results) <- unique_ids
```

# GeneID Conversions for Object

```{r}
#Converting EntrezIDs into Symbols
id_chunks <- split(unique_ids, ceiling(seq_along(unique_ids) / 10000))

for(i in seq_along(id_chunks)){
  chunk_map <- AnnotationDbi::mapIds(org_db, 
                      keys = id_chunks[[i]],
                      keytype = local_from,
                      column =   local_to,
                      multiVals = "first")
  #updating the inital final results
  final_results[names(chunk_map)] <- chunk_map
}

final_results <- unlist(final_results)
missing_ids <- names(final_results)[is.na(final_results)]
length(missing_ids)



if (length(missing_ids) > 0){
  
      layer2 <- getBM(
        attributes = c(biomart_from, biomart_to),
        filters = biomart_from,
        values = missing_ids,
        mart = useMart("ensembl", dataset = "hsapiens_gene_ensembl")
      )
      layer2
  if(nrow(layer2) > 0) {
    indices <- match(layer2[[biomart_from]], names(final_results))
    final_results[indices] <- layer2[[biomart_to]]
  }
}

missing_ids <- names(final_results)[is.na(final_results)]
length(missing_ids)
```

Is it possible to get the converted symbols into ensemble IDS in totality?

If this works then we can use multiple dbs on the symbols as a universal token

then convert them all back to Ensemble IDs

match their ENTREZID index back to the symbols then boom

we now have a totally converted list of ENTREZ –\> Symbol –\> ENSEMBL pipeline

Journal Note:

I'd rather have this project done and posted than trying to perfect the

Although it was a good try it's likely that those missing 1092X gene are ncRNA or tRNA and not protein coding GENEs

Important Note Down: Resulting dds object for bulk transcriptomic data of the second data set will not have all genes. 10,925 will be missing because the ENSEMBL IDS were not available.

If there is someone out there that can convert those remaining 10,925

```{r}
mapping_df <- data.frame(geneid = names(final_results),
                         ensembl_id = as.character(final_results))

final_tbl <- tbl %>% 
  left_join(., mapping_df, by = "geneid") %>%
  dplyr::relocate(ensembl_id, .after = "geneid")

#Slicing Rows with any NAs within the Data
final_tbl <- final_tbl[complete.cases(final_tbl),]


final_tbl <- final_tbl %>%
    group_by(ensembl_id) %>%
    summarise(across(where(is.numeric), sum)) %>%
    column_to_rownames(var = "ensembl_id")

#final edits of the metadata
meta_227116$samples <- paste0(meta_227116$samples, "_", meta_227116$condition)

meta_227116 <- meta_227116 %>% dplyr::select(-c(title))
```

```{r}
covid19_ds2 <- make_dds(counts = final_tbl , metadata = meta_227116, ah_record = "AH89426")
```

```{r}
plot_total_counts(covid19_ds2)
plot_library_complexity(covid19_ds2)
plot_gene_detection(covid19_ds2)
plot_biotypes(covid19_ds2)
```

# Reading in GSE161731

Illumina NovaSeq 6000 (Homo sapiens)

46 individuals with PCR positive, sympomatic COVID-19, 14 were sampled at multiple time points

(early ≤10 days, middle 11-21 days, late \>21 days)

**Categorical severity of Healthy and COVID-19 patients**

59 = acute respiratory infection

17 = Influenza

20 = bacterial pneumonia

19 = healthy controls

```{r}
gse_161731 <- getGEO("GSE161731", GSEMatrix = TRUE)
gse_161731 <- gse_161731[[1]]

expr_161731 <- exprs(gse_161731)
meta_161731<- pData(gse_161731)

head(expr_161731)
head(meta_161731)

getGEOSuppFiles("GSE161731")
```

### Aesthetic mapping of metadata

```{r}
meta_161731 <- meta_161731 %>%
  mutate(characteristics_ch1.1 = gsub("age: ", "", characteristics_ch1.1)) %>%
  mutate(age = as.integer(characteristics_ch1.1)) %>%
  
  # Gender
  mutate(characteristics_ch1.2 = gsub("gender: ", "", characteristics_ch1.2)) %>%
  rename(gender = characteristics_ch1.2) %>%
  
  # Race
  mutate(characteristics_ch1.3 = gsub("race: ", "", characteristics_ch1.3)) %>%
  rename(race = characteristics_ch1.3) %>%
  
  # Cohort
  mutate(characteristics_ch1.4 = gsub("cohort: ", "", characteristics_ch1.4)) %>%
  rename(cohort = characteristics_ch1.4) %>%
  
  # Hospitalized
  mutate(characteristics_ch1.5 = gsub("hospitalized: ", "", characteristics_ch1.5)) %>%
  rename(hospitalized = characteristics_ch1.5) %>%
  
  # Tissue
  mutate(characteristics_ch1.6 = gsub("tissue: ", "", characteristics_ch1.6)) %>%
  rename(tissue = characteristics_ch1.6) %>%
  
  # Batch (Converted to Integer)
  mutate(characteristics_ch1.7 = gsub("batch: ", "", characteristics_ch1.7)) %>%
  mutate(batch = as.integer(characteristics_ch1.7)) %>%
  select(geo_accession, gender, race, cohort)
meta_161731
```

```{r}
urld_ds3 <- "https://www.ncbi.nlm.nih.gov/geo/download/?format=file&type=rnaseq_counts"
path_ds3 <- paste(urld_ds3, "acc=GSE161731", "file=GSE161731_raw_counts_GRCh38.p13_NCBI.tsv.gz", sep="&");
tbl_ds3 <- as.data.frame(as.matrix(data.table::fread(path_ds3, header=T, colClasses="integer"), rownames=1))



```

### Mapping Sample Name to Condition

```{r}
#Mapping sample to the Condition
mapping <- c(meta_161731$cohort)
names(mapping) <- meta_161731$geo_accession

column_conditions <- mapping[colnames(tbl_ds3)]

new_names <- paste(colnames(tbl_ds3), column_conditions, sep = "_")

colnames(tbl_ds3) <- new_names

tbl_ds3 <- rownames_to_column(tbl_ds3, "geneid")
```

```{r}
# initial args
from_type <- "ENTREZ"

to_type <- "ENSEMBL"

org_db <- org.Hs.eg.db

# Harmoizing between each of the available annotation functions
harmonizer <- list(
    ENTREZ = list(local = "ENTREZID", biomart = "entrezgene_id", gconvert = "ENTREZGENE_ACC"),
    ENSEMBL = list(local = "ENSEMBL",  biomart = "ensembl_gene_id", gconvert = "ENSG"),
    SYMBOL  = list(local = "SYMBOL",   biomart = "external_gene_name",
                   gconvert = "SYMBOL")
  )

local_from   <- harmonizer[[from_type]]$local
local_to     <- harmonizer[[to_type]]$local
  
biomart_from <- harmonizer[[from_type]]$biomart
biomart_to   <- harmonizer[[to_type]]$biomart




unique_ids <- unique(tbl_ds3$geneid)


#Intialize final results
unique_ids_len <- length(unique_ids)

final_results <- c(replicate(unique_ids_len, NA))

names(final_results) <- unique_ids

#Converting EntrezIDs into Symbols
id_chunks <- split(unique_ids, ceiling(seq_along(unique_ids) / 10000))

for(i in seq_along(id_chunks)){
  chunk_map <- AnnotationDbi::mapIds(org_db, 
                      keys = id_chunks[[i]],
                      keytype = local_from,
                      column =   local_to,
                      multiVals = "first")
  #updating the inital final results
  final_results[names(chunk_map)] <- chunk_map
}



#Converting EntrezIDs into Symbols
id_chunks <- split(unique_ids, ceiling(seq_along(unique_ids) / 10000))

for(i in seq_along(id_chunks)){
  chunk_map <- AnnotationDbi::mapIds(org_db, 
                      keys = id_chunks[[i]],
                      keytype = local_from,
                      column =   local_to,
                      multiVals = "first")
  #updating the inital final results
  final_results[names(chunk_map)] <- chunk_map
}

final_results <- unlist(final_results)
missing_ids <- names(final_results)[is.na(final_results)]
length(missing_ids)



if (length(missing_ids) > 0){
  
      layer2 <- getBM(
        attributes = c(biomart_from, biomart_to),
        filters = biomart_from,
        values = missing_ids,
        mart = useMart("ensembl", dataset = "hsapiens_gene_ensembl")
      )
      layer2
  if(nrow(layer2) > 0) {
    indices <- match(layer2[[biomart_from]], names(final_results))
    final_results[indices] <- layer2[[biomart_to]]
  }
}

missing_ids <- names(final_results)[is.na(final_results)]
length(missing_ids)

mapping_df <- data.frame(geneid = names(final_results),
                         ensembl_id = as.character(final_results))

final_tbl_ds3 <- tbl_ds3 %>% 
  left_join(., mapping_df, by = "geneid") %>%
  dplyr::relocate(ensembl_id, .after = "geneid")

#Slicing Rows with any NAs within the Data
final_tbl_ds3 <- final_tbl_ds3[complete.cases(final_tbl_ds3),]


final_tbl_ds3 <- final_tbl_ds3 %>%
    group_by(ensembl_id) %>%
    summarise(across(where(is.numeric), sum)) %>%
    column_to_rownames(var = "ensembl_id")
```

```{r}
#final meta_data editing
meta_161731$samples <- paste0(meta_161731$geo_accession, "_", meta_161731$cohort)
meta_161731 <- meta_161731 %>% relocate(samples, .after = geo_accession) %>% select(samples, gender, race)
rownames(meta_161731) <- NULL
```

```{r}
covid19_ds3 <- make_dds(counts = final_tbl_ds3 , metadata = meta_161731, ah_record = "AH89426")
plot_total_counts(covid19_ds3)
plot_library_complexity(covid19_ds3)
plot_gene_detection(covid19_ds3)
plot_biotypes(covid19_ds3)
```
