---
title: "Meta_Analysis"
output: html_document
date: "2026-06-06"
---

```{r, include =FALSE}
library(tidyverse)
library(glue)
library(devtools)
library(vroom)
library(GEOquery)
library(biomaRt)
library(dplyr)
library(org.Hs.eg.db)
library(DESeq2)
library(clusterProfiler)
library(RNAseqQC)
library(edgeR)
library(ggvenn)
source("Project_1_function_lab.R")
```

# Loading Downstream Results from Intra and Inter Differential Experssion.

```{r}

full_path <- "/Users/gbolden/Desktop/SPG2026/BIDS_2026/Project_1"

load(glue::glue("{full_path}/downstream_results/covid19_ds1_analysis.RData"))

load(glue::glue("{full_path}/downstream_results/covid19_ds2_analysis.RData"))

load(glue::glue("{full_path}/downstream_results/covid19_ds3_analysis.RData"))

load(glue::glue("{full_path}/downstream_results/covid19_ds4_analysis.RData"))

load(glue::glue("{full_path}/downstream_results/covid19_ds5_analysis.RData"))

load("/Users/gbolden/Desktop/SPG2026/BIDS_2026/Project_1/downstream_results/covid_ds12345_integrated.RData")
```

# Summarizing Total DEGs per Dataset

```{r}

deg_summary <- function(deg_comps){
  summary_table <- list()
  
  for(comp in names(deg_comps)) {
  
    tbl <- deg_comps[[comp]] %>%
      dplyr::count(color, name = comp) %>%
      tibble::column_to_rownames("color")
  
    tbl <- rbind(
      tbl,
      total_DEGs = colSums(tbl)
    )
    summary_table[[comp]] <- tbl
  }
  
  summary_table <- do.call(cbind, summary_table)
  
  return(summary_table)
}


all_summaries <- lapply(
  list(
    gse202805 = covid19_ds1_analysis$filtered_results,
    gse152641 = covid19_ds2_analysis$filtered_results,
    gse161731 = covid19_ds3_analysis$filtered_results,
    gse227116 = covid19_ds4_analysis$filtered_results,
    gse152418 = covid19_ds5_analysis$filtered_results,
    dsint = covid19_ds124_analysis$filtered_results
  ),
  deg_summary
)
```

```{r}
transposed_summaries <- lapply(all_summaries, t)
transposed_summaries <- lapply(transposed_summaries, as.data.frame)

long_tbl <- imap_dfr(
  transposed_summaries,
  ~ .x %>%
      rownames_to_column("comparison") %>%
      mutate(dataset = .y)
)
long_tbl


long_tbl %>%
  ggplot(aes(x = comparison, y = total_DEGs, fill = dataset)) +
  geom_col() +
  theme(
    axis.text.x = element_text(angle = 60, hjust = 1)
  ) +
  labs(
    title = "Total DEGs Across Comparisons per Dataset",
    x = "Comparison",
    y = "Total DEGs"
  )


long_tbl %>%
  ggplot(aes(x = comparison, y = Downregulated, fill = dataset)) +
  geom_col() +
  theme(
    axis.text.x = element_text(angle = 60, hjust = 1)
  ) +
  labs(
    title = "Uppregulated DEGs Across Comparisons per Dataset",
    x = "Comparison",
    y = "Total DEGs"
  )

long_tbl %>%
  ggplot +
  aes(x = comparison, y = Upregulated, fill = dataset) +
 geom_col() +
  theme(
    axis.text.x = element_text(angle = 60, hjust = 1)
  ) +
  labs(
    title = "Downregulated DEGs Across Comparisons per Dataset",
    x = "Comparison",
    y = "Total DEGs"
  )

```

# Total DEGs per Dataset

```{r}
plot_tbl <- long_tbl %>%
  pivot_longer(
    cols = c(Downregulated, Upregulated, total_DEGs),
    names_to = "DEG_type",
    values_to = "count"
  )
plot_tbl
```

```{r}
plot_tbl %>%
  filter(DEG_type != "total_DEGs") %>%
  ggplot(aes(x = comparison, y = count, fill = DEG_type)) +
  geom_col() +
  scale_fill_manual(values = c(
    "Upregulated" = "red",
    "Downregulated" = "blue"
  )) +
  facet_wrap(~ dataset) +
  theme(
    axis.text.x = element_text(angle = 60, hjust = 1, vjust = 1, size = 10, face = "bold"),
    strip.text = element_text(face = "bold", size = 10),
    plot.title = element_text(hjust = 0.5, size = 20, face = "bold")
  ) +
  labs(title = "DEGs per Comparison by Dataset",
       y = "Number of DEGs")
```

# Total DEG and ORA Overlaps

if the comparison exists within those dataset

```{r}
deg_results_list <- list(
    gse202805 = covid19_ds1_analysis$filtered_results,
    gse152641 = covid19_ds2_analysis$filtered_results,
    gse161731 = covid19_ds3_analysis$filtered_results,
    gse227116 = covid19_ds4_analysis$filtered_results,
    gse152418 = covid19_ds5_analysis$filtered_results,
    dsint = covid19_ds124_analysis$filtered_results
)

# initialize output list
all_degs <- vector("list", length(deg_results_list))
names(all_degs) <- c("GSE202805", 
                     "GSE152641",
                     "GSE161731",
                     "GSE227116",
                     "GSE152418", 
                     "Integrated")

# loop through each dataset
for (j in seq_along(deg_results_list)) {
  
  current_dataset <- deg_results_list[[j]]
  current_genes <- c()
  
  # loop through each DEG table in the dataset
  for (i in seq_along(current_dataset)) {
    
    current_genes <- c(
      current_genes,
      current_dataset[[i]]$geneid
    )
  }
  
  # optionally remove duplicates
  all_degs[[j]] <- unique(current_genes)
}


ggvenn(all_degs[names(all_degs) != "Integrated"],
       show_percentage = FALSE,
       text_size = 5)

#Finding the most frequencted DEGS for all Dataset
core_genes <- Reduce(intersect, all_degs)

length(core_genes)
head(core_genes)

gene_freq <- table(unlist(all_degs))

shared_5plus <- names(gene_freq[gene_freq >= 5])
length(shared_5plus)
```

What are those 60 DEGs

```{r}

intersection_assignments <- all_degs %>%
  enframe(name = "set", value = "geneid") %>%
  unnest(geneid) %>%
  group_by(geneid) %>%
  summarize(Intersection = paste(sort(set), collapse = " & "))

# Split into a clean list of genes per intersection group
genes_by_intersection <- split(intersection_assignments$geneid, intersection_assignments$Intersection)

genes_by_intersection$`dsint & gse152418 & gse152641 & gse161731 & gse202805`
genes_by_intersection$`dsint & gse152418 & gse152641 & gse202805`
genes_by_intersection$`dsint & gse152418 & gse152641 & gse161731`

```

### Overlapping Overrepresented Terms

```{r}
ora_desc_collapse_fxn <- function(unpacked_results){
  ora_desc_list <- vector("list", length(unpacked_results))
  names(ora_desc_list) <- names(unpacked_results)
  
  for(i in seq_along(unpacked_results)) {
    
    # for each comparison within the unpacked results
    current_comparison <- unpacked_results[[i]]
    
    ora_descriptions <- c()
    
    # for each table within that specific comparison
    for(j in seq_along(current_comparison)) {
      
      ora_descriptions <- c(
        ora_descriptions,
        current_comparison[[j]]$gene_desc.x
      )
    }
    
    ora_desc_list[[i]] <- unique(ora_descriptions)
  }
  return(ora_desc_list)
}
```

```{r}
analysis_list <- list(
  GSE202805  = covid19_ds1_analysis,
  GSE152641  = covid19_ds2_analysis,
  GSE161731  = covid19_ds3_analysis,
  GSE227116  = covid19_ds4_analysis,
  GSE152418  = covid19_ds5_analysis,
  Integrated = covid19_ds124_analysis
)

all_ora_terms <- lapply(analysis_list, function(x) {
  c(
    up   = ora_desc_collapse_fxn(x$unpacked_results_up),
    down = ora_desc_collapse_fxn(x$unpacked_results_down)
  ) %>% unlist()
})

ggvenn(all_ora_terms[1:5])


#Finding the most frequencted DEGS for all Dataset
core_ora_terms <- Reduce(intersect, all_ora_terms[1:5])

length(core_ora_terms)
head(core_ora_terms)

gene_freq <- table(unlist(all_degs))

shared_5plus <- names(gene_freq[gene_freq >= 5])
length(shared_5plus)
```

### Unpacking the universally shared ORAs

```{r}
#Finding the most frequencted DEGS for all Dataset
core_ora_terms <- Reduce(intersect, all_ora_terms[1:5])

length(core_ora_terms)
core_ora_terms

gene_freq <- table(unlist(core_ora_terms))

shared_5plus <- names(gene_freq[core_ora_terms >= 5])
length(shared_5plus)
```

# **Healthy-Covid19 Comparison Across Dataset**

Comparison Upregulated and Down-Regulated DEGs across the Datasets for the most common comparison: Healthy-Covid19.

```{r}
library(UpSetR)
deg_results_list <- list(
    gse202805 = covid19_ds1_analysis$filtered_results$`healthy-covid19`,
    gse152641 = covid19_ds2_analysis$filtered_results$`healthy-covid19`,
    gse161731 = covid19_ds3_analysis$filtered_results$`healthy-covid19`,
    gse152418 = covid19_ds5_analysis$filtered_results$`healthy-covid19`,
    dsint = covid19_ds124_analysis$filtered_results$`healthy-covid19`
)

#Nested Upregulated and Downregulated DEGs for each dataset with comparison health-covid19
up_reg_degs   <- list()
down_reg_degs <- list()
all_degs      <- list()

for(dataset in names(deg_results_list)){
  up_reg_degs[[dataset]]   <- list()
  down_reg_degs[[dataset]] <- list()
  all_degs[[dataset]]      <- list()
  
    
  deg_tbl <- deg_results_list[[dataset]]

  up_reg_degs[[dataset]] <-
      unique(deg_tbl$geneid[deg_tbl$log2foldchange > 1])

    down_reg_degs[[dataset]] <-
      unique(deg_tbl$geneid[deg_tbl$log2foldchange < -1])

  all_degs[[dataset]] <-
      unique(deg_tbl$geneid)
}

#Upset Plots of the DEGs
up_reg_degs_upsetdata <- UpSetR::fromList(up_reg_degs)
UpSetR::upset(
  up_reg_degs_upsetdata,
  order.by = "freq",
  nsets = length(up_reg_degs),
  sets.bar.color = "red",
  main.bar.color = "red",
  text.scale = 5
)

down_reg_degs_upsetdata <- UpSetR::fromList(down_reg_degs)
UpSetR::upset(
  down_reg_degs_upsetdata,
  order.by = "freq",
  nsets = length(down_reg_degs),
  sets.bar.color = "blue",
  main.bar.color = "blue",
  text.scale = 5
)
```

```{r}
#Venn Diagram of the Plots
all_degs_upsetdata <- UpSetR::fromList(all_degs)
upset(
  all_degs_upsetdata,
  order.by = "freq"
)

ggvenn(all_degs[names(all_degs) != "Integrated"],
       show_percentage = FALSE,
   text_size = 5)

#Finding the most frequencted DEGS for all Dataset
core_genes <- Reduce(intersect, all_degs)

length(core_genes)
head(core_genes)

gene_freq <- table(unlist(all_degs))

shared_5plus <- names(gene_freq[gene_freq >= 5])
length(shared_5plus)
```

```{r}

initial_table <- covid19_dsint

table_list <- covid19_ds124_analysis$filtered_results$`healthy-covid19`[covid19_ds124_analysis$filtered_results$`healthy-covid19`$geneid %in% core_genes, ]

table_list <- list(table_list)
names(table_list) <- 'healthy-covid19'
```

```{r}
heatmap_function(initial_table, table_list)
```

ORA DEG consolidation

```{r}
#Dataset1 DEG Unpacking
#DEGs of this specific terms unpacked for this dataset
ds1_ora_t1 <- covid19_ds1_analysis$unpacked_results_down$`healthy-covid19`$`immunoglobulin mediated immune response`

ds1_ora_t2 <- covid19_ds1_analysis$unpacked_results_down$`healthy-covid19`$`B cell mediated immunity`

ds1_ora_t3 <- covid19_ds1_analysis$unpacked_results_down$`healthy-covid19`$`antibacterial humoral response`


#Dataset2 DEG unpacking
ds2_ora_t1 <- covid19_ds2_analysis$unpacked_results_down$`healthy-covid19`$`defense response to virus`

#Dataset3 DEG unpacking
ds3_ora_t1 <- covid19_ds3_analysis$unpacked_results_down$`healthy-covid19`$`B cell mediated immunity`

ds3_ora_t2 <- covid19_ds3_analysis$unpacked_results_down$`healthy-covid19`$`immunoglobulin mediated immune response`


#Dataset5 DEG unpacking
ds5_ora_t1 <- covid19_ds5_analysis$unpacked_results_down$`healthy-covid19`$`immunoglobulin mediated immune response`

ds5_ora_t2 <- covid19_ds5_analysis$unpacked_results_down$`healthy-covid19`$`B cell mediated immunity`


ora_gene_lists <- list(
  ds1_ora_t1 = ds1_ora_t1$geneid,
  ds1_ora_t2 = ds1_ora_t2$geneid,
  ds1_ora_t3 = ds1_ora_t3$geneid,
  ds2_ora_t1 = ds2_ora_t1$geneid,
  ds3_ora_t1 = ds3_ora_t1$geneid,
  ds3_ora_t2 = ds3_ora_t2$geneid,
  ds5_ora_t1 = ds5_ora_t1$geneid,
  ds5_ora_t2 = ds5_ora_t2$geneid
)

ora_upset_data <- UpSetR::fromList(ora_gene_lists)

upset(
  ora_upset_data,
  nsets = 8,
  nintersects = 100,
  order.by = "freq"
)

library(ComplexHeatmap)

library(tidyverse)

intersection_assignments <- ora_gene_lists %>%
  enframe(name = "set", value = "geneid") %>%
  unnest(geneid) %>%
  group_by(geneid) %>%
  summarize(Intersection = paste(sort(set), collapse = " & "))

# Split into a clean list of genes per intersection group
genes_by_intersection <- split(intersection_assignments$geneid, intersection_assignments$Intersection)

genes_by_intersection
```

If we consolidated all the DEGs that were aligned under immunological over-represented terms.

```{r}
df_ora_filtered <- unique(bind_rows(
  ds1_ora_t1,
  ds1_ora_t2,
  ds1_ora_t3,
  ds2_ora_t1,
  ds3_ora_t1,
  ds3_ora_t2,
  ds5_ora_t1,
  ds5_ora_t2
), by = "gene_id")
df_ora_filtered <- list(df_ora_filtered)
names(df_ora_filtered) <- 'healthy-covid19'


heatmap_function(covid19_ds1, df_ora_filtered) #112 DEGs
heatmap_function(covid19_ds2, df_ora_filtered)
heatmap_function(covid19_ds3, df_ora_filtered)
heatmap_function(covid19_ds5, df_ora_filtered)
heatmap_function(covid19_dsint, df_ora_filtered)
```

```{r}
#DownRegulated ORA Terms within the healthy-covid19
covid19_ds1_analysis$ora_down_raw$enrichgo_visuals$`healthy-covid19`$ora_pvalue_ranked
covid19_ds2_analysis$ora_down_raw$enrichgo_visuals$`healthy-covid19`$ora_pvalue_ranked
covid19_ds3_analysis$ora_down_raw$enrichgo_visuals$`healthy-covid19`$ora_pvalue_ranked
covid19_ds5_analysis$ora_down_raw$enrichgo_visuals$`healthy-covid19`$ora_pvalue_ranked
covid19_ds124_analysis$ora_down_raw$enrichgo_visuals$`healthy-covid19`$ora_pvalue_ranked

#UpRegulated ORA Terms within the healthy-covid19
covid19_ds1_analysis$ora_up_raw$enrichgo_visuals$`healthy-covid19`$ora_pvalue_ranked
covid19_ds2_analysis$ora_up_raw$enrichgo_visuals$`healthy-covid19`$ora_pvalue_ranked
covid19_ds3_analysis$ora_up_raw$enrichgo_visuals$`healthy-covid19`$ora_pvalue_ranked
covid19_ds5_analysis$ora_up_raw$enrichgo_visuals$`healthy-covid19`$ora_pvalue_ranked
covid19_ds124_analysis$ora_up_raw$enrichgo_visuals$`healthy-covid19`$ora_pvalue_ranked
```

```{r}
covid19_ds124_analysis$unpacked_results_up$`healthy-covid19`$`T cell differentiation`$geneid
```

Dissecting what DEGs makeupm those ORAs

```{r}
covid19_ds2_analysis$unpacked_results_down$`healthy-covid19`$

```
