---
title: "Intra_Differential_Expression"
output: html_document
---

Read-In Libraries

```{r, include =FALSE}
library(tidyverse)
library(glue)
library(devtools)
library(vroom)
library(GEOquery)
library(biomaRt)
library(dplyr)
library(org.Hs.eg.db)
library(DESeq2)
library(clusterProfiler)
library(RNAseqQC)
library(edgeR)
source("Project_1_function_lab.R")
```

```{r}
covid19_ds1 <- readRDS("./post_processed_data/gse202805_covid19_ds1.rds")
covid19_ds2 <- readRDS("./post_processed_data/gse152641_covid19_ds2.rds")
covid19_ds4 <- readRDS("./post_processed_data/gse227116_covid19_ds4.rds")
```

<https://olvtools.com/en/documents/edger>

## Perform EdgeR Differential Expression and plot a volcano plot for each condition comparison within the DDS object

```{r}

#Arguments
dds_object <- covid19_ds1
condition_col = "condition"
ref_group_name = "healthy"


#Performs inital differential expression on your DDS Object's raw counts dataframe
# The use give the condition column from the metadata to make the test parameters
# and the control group
results_dds1 <- edgeR_diffexp(covid19_ds1, condition_col,  ref_group_name)
results_dds2 <- edgeR_diffexp(covid19_ds2, condition_col, "Healthy")
results_dds4 <- edgeR_diffexp(covid19_ds4, condition_col, ref_group_name)
```

```{r}

# function starts here
#create a list of 


initial_table <- covid19_ds1


diffexp_results <- vector("list", length(results_dds1))
names(diffexp_results) <- names(results_dds1)

# for each comparison
for (i in seq_along(results_dds1)) {

  volcano_res <- volcano_plot(
    results_dds1[[i]],
    log2fc_thresh = 1,
    p.val_thresh = 0.05,
    lower_xlim = -10,
    upper_xlim = 10,
    step = 5
  )
  
  
  diffexp_results[[i]] <- list(

    data = volcano_res[[1]],

    vol_plot = volcano_res[[2]] +
      labs(title = glue::glue(
        "Volcano Plot: {names(results_dds1)[i]}"
      ))

  )
}
heatmap_visuals <- heatmap_function(initial_table, filtered_results)
```

```{r}
filtered_results <- lapply(diffexp_results , function(x) x$data)
for(table in seq_along(filtered_results)){
  
}
results_vol_plots <- lapply(diffexp_results, function(x) x$vol_plot)


#perform an EnrichGO ORA
#ORA only takes in significant DEGs Answers the question:
#Which pathways are overrepresented in my list of significant genes? 

ora_output_up <- ora_enrichgo(
  filtered_results,
  direction = "up",
  log2fc_thresh = 1,
  pval_thresh = 0.05,
  top_terms = 15
)

ora_output_down <- ora_enrichgo(
  filtered_results,
  direction = "down",
  log2fc_thresh = 1,
  pval_thresh = 0.05,
  top_terms = 15
)
```

```{r}
diffexp_results
filtered_results
heatmap_visuals
ora_output_up
ora_output_down
```

# Inspect your enirchGO ORA results

Unpacking ORA from GO and align to your differential expression results for the input condition

ORA only takes in significant DEGs Answers the question:

**Which pathways are overrepresented in my list of significant genes?**

```{r}
enrichgo_unpack_ver3 <- function(
                                    # Pass the whole list so we can grab comparisons
                                    initial_table_list = filtered_results,
                                    key = ora_output_down$enrichgo_results,
                                    pval_adj_threshold = 0.001,
                                    
                                    # Assume user does not want to convert geneids
                                    convert = FALSE,
                                    from_type = "ensembl",
                                    to_type = "symbol",
                                    ensembl_dataset = "hsapiens_gene_ensembl"){
  require(tibble)
  require(dplyr)
  
  ### helper functionf or converting geneidA to geneidB
  ### user
  gene_id_converter <- function(vector, from_type, to_type, ensembl_dataset){
    require(biomaRt)
    # geneid pulling for correct attribute
    id_map <- c(
      ensembl = "ensembl_gene_id",
      entrez  = "entrezgene_id",
      symbol  = "external_gene_name"
    )
    # p
    from_attr <- id_map[[from_type]]
    to_attr <-id_map[[to_type]]
    
    ensembl <- useMart(biomart = "ensembl", dataset = ensembl_dataset)
    
    return_df <- getBM( attributes = c(from_attr, to_attr), filters = from_attr, values = vector, mart = ensembl
    )
    return(return_df)
    
  }
  
  ###creates a helper function to unlist and strsplits the enriched Gene IDs per term
  unlist_converter <- function(str) {
    if (grepl("^ENSG|\\d", str)){
      bruh <- as.character(unlist(strsplit(str, "/")))
    }else{
      bruh <- as.integer(unlist(strsplit(str, "/")))
    }
    return(bruh)
  }

  
  # FIX 2: Corrected the spelling typo here
  unpacking_results <- list()
  
  # FIX 1: Loop over names(key) so 'comparison' is a valid character string identifier
  for (comparison in names(key)) {
    
    # Safeguard check: If the comparison doesn't exist in the cluster results, skip
    if (is.null(key[[comparison]])) next
    
    cluster <- tibble::as_tibble(key[[comparison]]@result) %>%
      dplyr::mutate(Description = as.character(Description))
    
    if (!is.null(pval_adj_threshold)) {
      significant_clusters <- cluster %>%
        dplyr::filter(p.adjust < pval_adj_threshold)
    } else {
      significant_clusters <- cluster
    }
    
    labeled_gene_lists <- list()
    
    # FIX 3: Pull the specific comparison dataframe from your filtered results list
    current_initial_table <- initial_table_list[[comparison]] %>%
      dplyr::mutate(geneid = as.character(geneid))
    
    # If there are no significant clusters, jump to saving the empty list or next comparison
    if (nrow(significant_clusters) == 0) {
      unpacking_results[[comparison]] <- labeled_gene_lists
      next
    }
    
    for (i in seq_len(nrow(significant_clusters))) {
      
      go_term <- significant_clusters$Description[i]
      
      # Extract genes for current GO term
      genes <- unlist_converter(significant_clusters$geneID[i])
      
      pathway_tibble <- tibble::tibble(
        gene_desc = rep(go_term, length(genes)), 
        geneid = as.character(genes)
      )
      
      if (convert == FALSE) {
        pathway_genenames <- pathway_tibble
      } else {
        pathway_genenames <- gene_id_converter(
          pathway_tibble$geneid,
          from_type = from_type,
          to_type = to_type,
          ensembl_dataset = ensembl_dataset
        )
        
        new_colname <- paste0(to_type, "_geneid")
        
        # FIX 4: Cleaner, bulletproof column renaming by position
        colnames(pathway_genenames)[1] <- "geneid"
        colnames(pathway_genenames)[2] <- new_colname
      }
      
      # Join converted names
      final_path_tibble <- dplyr::inner_join(pathway_tibble, pathway_genenames, by = "geneid") %>%
        dplyr::mutate(geneid = as.character(geneid))
      
      # Join fold changes / statistics using our dynamically updated initial table
      final_path_fc_tibble <- dplyr::inner_join(final_path_tibble, current_initial_table, by = "geneid")
      
      # Relocate converted gene column
      if (convert == TRUE) {
        fpfc_tibble <- final_path_fc_tibble %>% 
          dplyr::relocate(dplyr::all_of(new_colname), .after = geneid)
      } else {
        fpfc_tibble <- final_path_fc_tibble
      }
      
      # Store result under the GO Term name
      labeled_gene_lists[[go_term]] <- fpfc_tibble
    }
    
    # Store the list of terms into the master list under the Comparison name
    unpacking_results[[comparison]] <- labeled_gene_lists
  }
  return(unpacking_results)
}
```

```{r}
# making a data table for each Description Term that is consider statistically significant within the context of the ORA

initial_tables <- filtered_results

key <- ora_output_down$enrichgo_results
  
pval_adj_threshold <- 0.001
#assume user does not want to convert geneids
convert = FALSE

from_type = "ensembl"

to_type = "symbol"

ensembl_dataset <- "hsapiens_gene_ensembl"

unpacked_results <-  enrichgo_unpack_ver3(initial_table_list = initial_tables,
                     key = key,
                     pval_adj_threshold = 0.001,
                     convert = FALSE,
                     from_type = "ensembl",
                     to_type = "symbol",
                     ensembl_dataset = "hsapiens_gene_ensembl")
```

```{r}
unpacked_results
```
