---
title: "Project_1_Analysis"
output: html_document
---

# Library Read-In

```{r}
library(tidyverse)
library(devtools)
library(vroom)
library(GEOquery)
library(biomaRt)
library(dplyr)
library(org.Hs.eg.db)
library(DESeq2)
library(clusterProfiler)
library(RNAseqQC)
```

# Reading in the files

### GSE202805 Info

tissue: pbmc

**80 Samples:**

**Platform:** Illumina HiSeq 4000 (Homo sapiens)

**Case Groups:** 4 acute-Mild, 6 Acute Moderate, 32 Acute-Severe, 28 Covalescent

**Control Groups:** 10 healthy controls

### **GSE227116 Info**

tissue: pbmc

65 blood Samples recoved form 46 COVID-19 convalescent patients and 10 healthy donors were analyzed by RNA- Sequencing

**Platform:** Illumina HiSeq 2500 (Homo sapiens)

10 healthy donors, 65 blood samples.

### GSE161731 Info

tissue: pbmc

Illumina NovaSeq 6000 (Homo sapiens)

46 individuals with PCR positive, sympomatic COVID-19, 14 were sampled at multiple time points

(early ≤10 days, middle 11-21 days, late \>21 days)

**Categorical severity of Healthy and COVID-19 patients**

59 = acute respiratory infection

17 = Influenza

20 = bacterial pneumonia

19 = healthy controls

```{r}
covid19_ds1 <- readRDS("./post_processed_data/gse202805_covid19_ds1.rds")
covid19_ds2 <- readRDS("./post_processed_data/gse227116_covid19_ds2.rds")
covid19_ds3 <- readRDS("./post_processed_data/gse161731_covid19_ds3.rds")
```

#### QC Plots on raw count matrix

typing a qc function plain without arguments gives you the actual ggplot 2 function: neat

**Library Size:** the total number of counts for each sample

Ideally we should see a level distribution between each of samples within the same level of magnitude.

## Library Size

Size distribution is typical and normal for all three. However DS2 and DS3 are right skewed with out of range samples. Outlier samples that exceed the magnitude of the second IQR can indicate that there was PCR Over amplification and too many duplicate reads. If we samples low, then those have low counts. Then those have around 5 to 10 million reads and are usually excluded. DS3 has a few, but overall all three dataset should that there that most the samples are within range to avoid batch effects.

```{r}
plot_total_counts(covid19_ds1) +
  labs(title = "gse202805 / COVID-19 DS1 library sizes") 
  
  
plot_total_counts(covid19_ds2) + 
  labs(title = "gse227116 / COVID-19 DS2 library sizes")
plot_total_counts(covid19_ds3) + 
  labs(title = "gse161731 / COVID-19 DS3 library sizes")
```

## Library Complexity

Library Complexity: the fraction of counts that are taken up by what fraction of genes.

This helps us to assess transcriptional diversity. Ideally, a high quality library should have read distributed across a broad range of gene rather than being dominated by a select few

GSE20285:

Starting with major outlier on the left, this shows us that 50% of counts within that sample contains less than 0.01.% of the genes within the homo-sapien reference genome used fro alignment. This what is known as low complexity sample/library. This would mean that very little known genes probably around 100 out of thousands are being seen within the total amounts of reads within that sample.

Our main cluster of libraries: Looks like we have a healthy amount of healthy amount of variation across all libraries. For almost all of the samples, 50% of the counts account for the first 10% of the reads and flatten out around the 50% of genes at 95% of the counts within the majority of the libraries.

A general rule for grading library complexity is to find the intersection at 50% of the reads counted.

```{r}
plot_library_complexity(covid19_ds1) +
  labs(title = "gse202805 / COVID-19 DS1 library complexity") 
  
plot_library_complexity(covid19_ds2) + 
  labs(title = "gse227116 / COVID-19 DS2 library complexity")

plot_library_complexity(covid19_ds3) + 
  labs(title = "gse161731 / COVID-19 DS3 library complexity")
```

**Library complexity is only one component of sequencing saturation. It is also dependent on the sequencing depth: the total number of reads the machine produced per sample**

This is different from **sequencing saturation:** the measure of the fraction of library complexity that was sequenced in a given experiment.

Low Saturation: You have unique molecules in your sample that sequencer hasn't yet seen. Adding more reads will increase the number of genes you detect

High Saturation: You have already sequenced almost ever molecule in the library multiple time. Adding more reads will only produce duplicates

Your Saturation limit can also vary depending on the cell types you're profiling (<https://kb.10xgenomics.com/s/article/115005062366-What-is-sequencing-saturation>)

## **Gene Detection**

```{r}
plot_gene_detection
plot_gene_detection(covid19_ds1) +
  labs(title = "gse202805 / COVID-19 DS1 gene detection") 
  
plot_gene_detection(covid19_ds2) + 
  labs(title = "gse227116 / COVID-19 DS2 gene detection")

plot_gene_detection(covid19_ds3) + 
  labs(title = "gse161731 / COVID-19 DS3 gene detection")
```

## **Clustering Using Internal DDS functions**

Euclidean Distance

```{r}
assays(covid19_ds1) <- assays(covid19_ds1)[c("var_stable", "counts", "log_counts")]
```

```{r}
plot_sample_clustering(covid19_ds1, anno_vars = c("condition"), distance = "euclidean")
```

Plotting the First 2 PCs

```{r}
plot_pca(covid19_ds1, PC_x = 1, PC_y = 2, color_by = "condition")
```

**Skree Plot to tell which PC Captures the most variability between between each sample**

Extacting top variable gene per sample

```{r}
# Setting Variance Stable Counts as Priority Assay
vst_mat <- assay(covid19_ds1, "var_stable")
#Select top variable gene
top_var_genes <- head(order(rowVars(vst_mat), decreasing = TRUE), 2000)
vst_top <- vst_mat[top_var_genes, ]
# Run PCA via R
# Remember PCA is taking the right unitary martrix (VT) (feature space) of SVD
# VT tells us the principal directions of the data set
# multiply the the original matrix by the right unitary which is equivalent to the
# U time signma (left unitary) * Singular values (Size of the PCs)
pca_res <- prcomp(t(vst_top))



```

**Calculating Percentage of Variance for the Scree Plot**

```{r}
# Calculate percentage variance explained
var_explained <- pca_res$sdev^2 / sum(pca_res$sdev^2) * 100

# Create a data frame for plotting (first 10 PCs)
scree_data <- data.frame(
  PC = paste0("PC", 1:10),
  Variance = var_explained[1:10]
)

# Ensure PC is treated as a factor in order
scree_data$PC <- factor(scree_data$PC, levels = scree_data$PC)
```

Scree Plot ensures that we are plotting PCs that capture the most amount of variability within our data

```{r}
library(ggplot2)

ggplot(scree_data, aes(x = PC, y = Variance, group = 1)) +
  geom_line(color = "red", size = 1) +
  geom_point(color = "red", size = 2) +
  labs(title = "Scree Plot: Variance Explained per PC",
       x = "Principal Component",
       y = "Percentage of Variance (%)") +
  theme_minimal()
```

## Plotting PC loadings:

**Which of the genes are considered drivers of variability within each sample ?**

```{r}
pca_res <- plot_pca(covid19_ds1, PC_x = 1, PC_y = 2, color_by = "condition", show_plot = FALSE)
plot_loadings(pca_res, PC = 1, annotate_top_n = 5)
```

## UMAP:

Input the Left Unitary \* Singular Values

```{r}
dim(t(assay(covid19_ds2, "var_stable")))
```

```{r}
# 1. Run PCA
pca_res <- prcomp(t(assay(covid19_ds1, "var_stable")), rank. = 50)

# 2. Use the 'x' slot (the PC scores) as input for UMAP
# We only take the number of PCs identified by your scree plot (e.g., 1:15)
library(umap)
umap_out <- umap(pca_res$x[, 1:15])


umap_df <- data.frame(
   Sample_IDS = rownames(pca_res$x),
   UMAP_1 = umap_out$layout[, 1],
   UMAP_2 = umap_out$layout[, 2],
   Condition = as.factor(colData(covid19_ds1)$condition))

ggplot(umap_df, aes(x = UMAP_1, y = UMAP_2, color = Condition)) +
  geom_point(size = 3) +
  theme_minimal() +
  labs(title = "UMAP of COVID-19 Samples", subtitle = "Input: Top 15 PCs from VST Data")
```

```{r}
length(colData(covid19_ds1)$condition)
```

```{r}
length(covid19_ds1$condition)
```
