---
title: "Intra_Differential_Expression"
output: html_document
---

Read-In Libraries

```{r, include =FALSE}
library(tidyverse)
library(glue)
library(devtools)
library(vroom)
library(GEOquery)
library(biomaRt)
library(dplyr)
library(org.Hs.eg.db)
library(DESeq2)
library(clusterProfiler)
library(RNAseqQC)
library(edgeR)
source("Project_1_function_lab.R")
```

```{r}
covid19_ds1 <- readRDS("./post_processed_data/gse202805_covid19_ds1.rds")
covid19_ds2 <- readRDS("./post_processed_data/gse152641_covid19_ds2.rds")
covid19_ds4 <- readRDS("./post_processed_data/gse227116_covid19_ds4.rds")
```

```{r}
covid19_ds2@colData
```

<https://olvtools.com/en/documents/edger>

## Perform EdgeR Differential Expression and plot a volcano plot for each condition comparison within the DDS object

```{r}

#Arguments
dds_object <- covid19_ds1
condition_col = "condition"
ref_group_name = "healthy"


#Performs inital differential expression on your DDS Object's raw counts dataframe
# The use give the condition column from the metadata to make the test parameters
# and the control group
results_dds1 <- edgeR_diffexp(covid19_ds1, condition_col,  ref_group_name)
results_dds2 <- edgeR_diffexp(covid19_ds2, condition_col, "Healthy")
results_dds4 <- edgeR_diffexp(covid19_ds4, condition_col, ref_group_name)
```

```{r}

# function starts here
#create a list of 
volcano_results <- vector("list", length(results_dds1))
names(volcano_results) <- names(results_dds1)

# for each comparison made
for (i in seq_len(length(results_dds1))) {
  # use the volcano_plot_function
  # to filter through your differential expression results
  
  
  volcano_res <- volcano_plot(
    results_dds1[[i]],
    log2fc_thresh = 1,
    p.val_thresh = 0.05,
    lower_xlim = -10,
    upper_xlim = 10,
    step = 5
  )
  
  #saving_resulting plots and data
 volcano_results[[i]] <- list(
    
    data = volcano_res[[1]],
    
    vol_plot = volcano_res[[2]] +
      labs(title = glue("Volcano Plot: {names(results_dds1)[i]}"))
  )
}

initial_table <- covid19_ds1

filtered_results = lapply(volcano_results , function(x) x$data)

results_vol_plots <- lapply(volcano_results, function(x) x$plot)

heatmap_function(initial_table, filtered_results)


```

```{r}
volcano_results
```

Creating A Heatmap

For each condition comparison

Arrange for top 100 DEGs in within the differential expression matrix

Pull those genes from the count matrices from the regular raw_count matrix

level samples

transpose –\> scale –\> transpose –\> Heatmap

```{r}
library(colorRamp2)
library(ComplexHeatmap)

initial_table <- covid19_ds1

filtered_results = lapply(volcano_results , function(x) x$data)


heatmap_function <- function(initial_table, table_list) {
  
  require(colorRamp2)
  require(ComplexHeatmap)
  
  # Initialize the list to store heatmap objects
  heatmap_list <- list()
  
  # Extract the raw count matrix once outside the loop for efficiency
  # Assumes a SingleCellExperiment or SummarizedExperiment-like structure
  expr_raw <- as.matrix(initial_table@assays@data$counts)
  
  # Loop through the list of filtered differential expression results
  for (i in seq_along(table_list)) {
    
    # Grab the ith table
    filtered_diff_exp_results <- table_list[[i]]
    
    # Split the names of the ctrl-treatment comparison
    comparison_name <- names(table_list)[i]
    ctrl_treatment <- strsplit(comparison_name, "-")[[1]]
    #itemizing the names of the
    ctrl_name      <- ctrl_treatment[1]
    treatment_name <- ctrl_treatment[2]
    
    # Message tracking
    message(glue::glue("Processing: {ctrl_name} vs {treatment_name}"))
    
    # Pull the differentially expressed gene IDs
    diffexp_geneids <- as.character(filtered_diff_exp_results$geneid)
    
    # Filter the matrix for the target genes and matching sample columns
    # Note: Added anchors to ensure exact suffix matching
    col_pattern <- glue::glue("_{ctrl_name}$|_{treatment_name}$")
    matching_cols <- grepl(col_pattern, colnames(expr_raw))
    
    expr <- expr_raw[diffexp_geneids, matching_cols, drop = FALSE]
    
    # Scaling and z-score normalizing the counts matrix
    expr_z <- t(scale(t(expr)))
    expr_z[is.na(expr_z)] <- 0
    rownames(expr_z) <- rownames(expr)
    colnames(expr_z) <- colnames(expr)
    
    # Gene ID conversion
    conversion_table <- gene_id_converter(
      rownames(expr_z), 
      "ensembl", 
      "symbol", 
      "hsapiens_gene_ensembl"
    )
    
    gene_map <- setNames(
      conversion_table$external_gene_name,
      conversion_table$ensembl_gene_id
    )
    
    new_names <- gene_map[rownames(expr_z)]
    
    # Fallback to Ensembl ID if no symbol is found
    rownames(expr_z) <- ifelse(is.na(new_names), rownames(expr_z), new_names)
    
    # Clustering data (Optional: ComplexHeatmap does this natively if cluster_rows = TRUE, 
    # but kept if you intend to pass hclust objects explicitly)
    gene_dist <- dist(expr_z)
    gene_hclust <- hclust(gene_dist)
    
    # Define color palette
    col_fun <- colorRamp2(c(-2, 0, 2), c("blue", "white", "red"))
    
    # Generate Heatmap object
    p <- Heatmap(
      expr_z, 
      name            = "Z-score",
      col             = col_fun,
      cluster_columns = TRUE,     
      cluster_rows    = TRUE, # To use your manual cluster, change to: cluster_rows = gene_hclust     
      row_names_gp    = gpar(fontsize = 12),
      column_title    = glue::glue("{ctrl_name} - {treatment_name} Diff Exp Genes")
    )
    
    # Store the plot object into the initialized list
    heatmap_list[[comparison_name]] <- p
  }
  
  return(heatmap_list)
}
```

```{r}

diffexp_geneids <- as.character(c(filtered_diff_exp_results$geneid))

expr <- as.matrix(covid19_ds1@assays@data$counts)

expr <- expr[diffexp_geneids, grepl("_covid19$|_healthy$", colnames(expr)), drop=FALSE]



expr_z <- t(scale(t(as.matrix(expr))))
expr_z[is.na(expr_z)] <- 0
# filtering for columns that are specific to that assay and differential expression result
rownames(expr_z) <- rownames(expr)
colnames(expr_z) <- colnames(expr)

#clustering data
gene_dist <- dist(expr_z)
gene_hclust <- hclust(gene_dist)

col_fun = colorRamp2(c(-2, 0, 2), c("blue", "white", "red"))
p <- Heatmap(expr_z, 
               name = "Z-score",
               col = col_fun,
               cluster_columns = TRUE,     
               cluster_rows = TRUE,     
               row_names_gp = gpar(fontsize = 12))
p
```

# Perform an EnrichGO ORA

```{r}

enrichgo_results <- vector("list", length(filtered_results))
names(enrichgo_results) <- names(filtered_results)

# for each compar
for(i in seq_along(filtered_results)){
  enrichgo_results[[i]] <- clusterProfiler::enrichGO(
          gene = filtered_results[[i]]$geneid,
           OrgDb = 'org.Hs.eg.db',
           keyType = "ENSEMBL",
           ont = "BP",
           pAdjustMethod = "BH")
}
```

# Inspect your enirchGO results

```{r}
enrichgo_results$`healthy-convalescent`@result
```

Unpacking ORA from GO and align to your differential expression results for the input condition

```{r}
# making a data table for each Description Term that is consider statistically significant within the context of the ORA

initial_table <- filtered_results$`healthy-covid19`

key <- enrichgo_results$`healthy-covid19`
    
chosen_desc <- "immunoglobulin mediated immune response"

#assume user does not want to convert geneids
convert = TRUE

from_type = "ensembl"

to_type = "symbol"

ensembl_dataset <- "hsapiens_gene_ensembl"

unpacked_results <-  enrichgo_unpack_ver2(intial_table = intial_table,
                     key = key,
                     chosen_desc = chosen_desc,
                     conver = TRUE,
                     from_type = "ensembl",
                     to_type = "symbol",
                     ensembl_dataset = "hsapiens_gene_ensembl")
```

To do list for unpacked results function

1.  scale to a whole
